---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "Manipulating data.frames with the dplyr library"
output:
  learnr::tutorial:
    includes:
      before_body: !expr system.file(file.path("tutorials", "style.html"),package="rtrainer")
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 


```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```

## Overview of Pipes (|>)

Pipes allow function calls to be chained together, making the code more readable. It can be read as a sequence of steps in which the result of the preceding function is sent, implicitly, to the first argument of the following function:

```{r example_pipe, exercise=TRUE}
rnorm(100, mean=10) |> 
  mean() |> 
  round(digit=2) |> 
  as.character()
```


## The dplyr library

The `dplyr` library is part of the tidyverse package. It is designed for fast, readable, and concise data manipulation. It simplifies common data manipulation tasks. It focuses on a set of function to make data transformations intuitive. The input to `dplyr` functions are always a `data.frame` (or `tibble` which is very close) and the output is always a `data.frame`. Functions (e.g. `select()`, `mutate()`, `filter()`) will typically receive the names of the columns (without quotes) to operate on.

We will illustrate the use of dplyr with the iris dataset that provide various flower measurement over 3 species of iris. 

```{r loadiris, exercise=TRUE,}
library(dplyr)
data(iris) 
head(iris)
```


## The select() function

Select a set of columns. This is an alternative to the indexing function.

```{r select, exercise=TRUE, exercise.setup="loadiris"}
iris |> 
  select(Sepal.Length, Sepal.Width, Species) |> 
  head()
```

Or `select()` a column that `matches` a regular expression (here something that ends with and "h"). 

```{r match, exercise=TRUE, exercise.setup="select"}
iris |> 
  select(matches("h$")) |> 
  head()
```

## The filter() function

Select a set of rows. This is an alternative to the indexing function.

```{r filter, exercise=TRUE, exercise.setup="select"}
iris |> 
  filter(Sepal.Length > 6 & Sepal.Width > 3.5)
```

## The mutate() function

This allow to create new columns that are functions of existing variables/columns:

```{r mutate, exercise=TRUE, exercise.setup="select"}
iris |> 
  select(matches("al")) |> 
  mutate(Sepal.diff=Sepal.Length - Sepal.Width,
         Petal.diff=Petal.Length - Petal.Width) |>
  head()
```


## The summarise() function

It can be viewed as an alternative to the apply (with MARGIN=2). It will return a data.frame after applying an operation over all rows of user-defined columns:

```{r summarise, exercise=TRUE, exercise.setup="select"}
iris |> summarise(mean.Sepal.Length=mean(Sepal.Length), 
                  mean.Sepal.Width=mean(Sepal.Width))
```

## The groupby() function

All this functions naturally combine with the `group_by()`function allowing to group rows by a set of categories. This may be particularly interesting to compute a statistic by category.

```{r group_by, exercise=TRUE, exercise.setup="select"}
iris |> 
  group_by(Species) |> 
  summarise(mean.Sepal.Length=mean(Sepal.Length), 
            mean.Sepal.Width=mean(Sepal.Width))
```

## Note

The `dplyr` library is quite a vast subject. We provide only a few basics in this tutorial. We encourage users to consult the [Data Transformation](https://r4ds.hadley.nz/data-transform) section of the “R for Data Science” book for more information.

### End of the section

Thank you for following this tutorial.

