---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "The ggplot2 library (session 2)"
output:
  learnr::tutorial:
    includes:
      before_body: !expr system.file(file.path("tutorials", "style.html"),package="rtrainer")
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 

```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```

## Histograms

### Simple Histograms

Analyzing **distributions** is **a crucial step in data exploration** and understanding the characteristics of a dataset. With ggplot2, we can easily create plots such as **histograms, density plots, boxplots, or violin plots** to analyze distributions.

To create a histogram for the "price" variable in the "diamonds" dataset, we use the `geom_histogram()` function. For a histogram, the x-axis represents intervals (*bins*), and the y-axis represents the count of observations within these intervals. Therefore, only one variable needs to be provided to the `aes()` function.

```{r histogram, exercise=TRUE}
library(ggplot2)
data(diamonds)

ggplot(data = diamonds, aes(x = price)) + geom_histogram()

```

We can slightly improve this histogram by changing the labels (e.g title and axis names) and the colors of the bars.

```{r histogram_2, exercise=TRUE, exercise.setup="histogram"}

ggplot(data = diamonds, aes(x = price)) + 
  geom_histogram(color = "black", 
                 fill = "lightblue") +
  labs(title = "Histogramme de Price",
       x = "Prix",
       y = "Fréquence")

```

<div class="exo">

- The key parameter of a histogram is the interval size. Here, in the `geom_histogram` function, it is controlled by the `binwidth` argument (interval size) or `bins` (number of intervals). Create histograms with interval sizes of \$250, \$500, \$1000, and \$10000.

```{r bins, exercise=TRUE, exercise.setup="histogram"}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1000)
```

```{r bins-solution}
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 250, color = "black", fill = "lightblue")
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 500, color = "black", fill = "lightblue")
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1000, color = "black", fill = "lightblue")
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(binwidth = 10000, color = "black", fill = "lightblue")
```
</div>


### Overlapping Histograms

In the context of `geom_point()`, the **geometry is a point**, and a variable can be associated with the color of the points. For a **histogram**, it is the same concept, except the **geometry is a histogram**. If we use `cut` for the color, we will obtain histograms with different colors.

<div class="exo">

- Complete the following code ("___") so that the coloring of the bar (`fill`) is associated with the quality of the `cut` column (the quality of the cut) of the diamond dataset. 


```{r histocut, exercise=TRUE , exercise.setup="histogram"}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = price, ___)) +
            geom_histogram(binwidth = 1000, 
                           color = "white") +
            scale_fill_manual(values=col_palette)
print(p)
```

```{r histocut-solution}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = price, 
                                 fill=cut)) +
            geom_histogram(binwidth = 1000, 
                           color = "white") +
            scale_fill_manual(values=col_palette)
print(p)
```

```{r histocut-check}
gradethis::grade_result_strict(
  pass_if(~ any(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$fill)) != c("#21908CFF", "#3B528BFF", "#440154FF", "#5DC863FF", "#FDE725FF")) == TRUE)
)
```
</div>

**Here we see the limitation of the histogram for visualizing distributions**. In this case, another solution to analyze price distributions is to use density analysis. Later, we will also see that it is possible to create histogram panels (facets).

## Density Profiles

The density plot is another way to visualize the distribution of values in a numeric variable. It can be created using the `geom_density()` function. The `geom_density()` function displays a distribution model that aims to closely fit the histogram data. 

```{r adjust, exercise=TRUE, exercise.setup="histogram"}
ggplot(data = diamonds, aes(x = price)) + geom_density()
```

<div class="exo">

- The `geom_density()` displays a distribution model. The `adjust` parameter controls the way the curve is adjusted to the data. It can significantly influence the representation. Depending on the value of `adjust`, the model may tend to be over- or under-fitted.
 Recreate the previous density plot using the following values for `adjust`: 1, 3/4, 1/2, 1/4, and 1/8.


```{r vardens, exercise=TRUE, exercise.setup="histogram"}
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = ___)
```

```{r vardens-solution}
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = 1)
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = 3/4)
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = 1/2)
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = 1/4)
ggplot(data = diamonds, aes(x = price)) + geom_density(adjust = 1/8)
```

</div>

<div class="exo">

- Complete the code to produce a plot where the densities for the different cuts (`cut`) are represented with a specific border color for each. Apply your own colors.

```{r vardens_2, exercise=TRUE, exercise.setup="histogram"}
color_palette <- c("Ideal" = ___, 
                   "Premium" = ___, 
                   "Very Good" = ___, 
                   "Good" = ___, 
                   "Fair" = ___)

p <- ggplot(data = diamonds, aes(x = price, ___ )) + 
  geom_density(adjust = 1/2, linewidth=0.5) +
  scale_color_manual(values=color_palette)

print(p)
```

```{r vardens_2-solution}
color_palette <- c("Ideal" = "#A22200", 
                   "Premium" = "#0871A4", 
                   "Very Good" = "#00B850", 
                   "Good" = "#226666", 
                   "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = price, color=cut )) + 
  geom_density(adjust = 1/2, linewidth=0.5) +
  scale_color_manual(values=color_palette)
  
print(p)
```

```{r vardens_2-check}

gradethis::grade_result_strict(
  pass_if(~ any(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$colour)) != c("#21908CFF", "#3B528BFF", "#440154FF", "#5DC863FF", "#FDE725FF")) == TRUE)
)
```

</div>

## Boxplots and Violin Plots

Boxplots (**boxplots**) and violin plots (**violin plots**) can be used to represent distributions associated with a dataset. Below are some examples.

As always, to create our plot we will first:

- *(i)* Associate data (data.frame or tibble) with an object (`data=diamonds`).
- *(ii)* Define **the variables/columns of the data.frame to analyze** and pass them to the **aes()** function.

For a **boxplot** or **violin plot**, the **aesthetic variables** accepted are generally the same as for **geom_points**. However, a key difference is that the *x* variable must be a factor (indicating the names of the boxes on the *x* axis) and the *y* variable must be numeric (the values taken by each category).

```{r firstgg2, exercise=TRUE , exercise.setup="histogram" }
ggplot(data=diamonds, aes(x=cut, y=price)) + geom_boxplot()
```

<div class="exo">

- Since `geom_boxplot()` and `geom_violin()` share the same aesthetic elements, complete the following code. This code creates a violin plot where the colors of the violins correspond to the cut type (`cut`).

```{r violin, exercise=TRUE , exercise.setup="histogram" }
col_palette <- c("Ideal" = "#A222A0", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=___)) + 
  geom_violin(color=___) +
  scale_fill_manual(values=___)

print(p)
```

```{r violin-solution}
col_palette <- c("Ideal" = "#A222A0", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")
p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut)) + 
  geom_violin(color="black") +
  scale_fill_manual(values=col_palette)

print(p)
```

```{r violin-check}

gradethis::grade_result_strict(
  pass_if(~ p$labels$x == "cut"),
  pass_if(~ p$labels$y == "price"),
  pass_if(~ p$labels$fill == "cut"),
  pass_if(~ any(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$fill)) != c("#21908CFF", "#3B528BFF", "#440154FF", "#5DC863FF", "#FDE725FF"))),
  pass_if(~ class(p$layers[[1]]$geom)[1] == "GeomViolin")  
)
```

</div>

## Overlaying Graphical Elements

### Example with the Violin Plot

The underlying ggplot model allows for relatively easy layering of graphical elements.

Consider the previous plot, $p$, created with the `geom_violin()` geometry.


```{r violinprepare, exercise=TRUE, exercise.setup="histogram" }
library(ggplot2)
data(diamonds)
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut))  +
  geom_violin(color="black") +
  scale_fill_manual(values=col_palette)

p
```

You might want to visualize the points that were used to construct the violin model. One might think to simply add a `geom_points()` geometry. Nothing could be easier: just add a `geom_point()` layer using the addition operator (`+`). The aesthetics are inherited, so `x=cut` and `y=price` are naturally passed to `geom_point()`. One issue is that the points are aligned vertically...

```{r violinvspoints, exercise=TRUE, exercise.setup="violinprepare"}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

 p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut)) +
        geom_violin(color="black") +
        scale_fill_manual(values=col_palette) + 
        geom_point()
p
```

This can be resolved by using `geom_jitter()` (instead of `geom_point()`), which adds **a bit of noise to the x-axis and y-axis**. As we only wants some "jittering" on the x axis we will indicate `height=0` (*i.e* no jittering on the y-axis).

```{r violinvsjitter, exercise=TRUE, exercise.setup="violinprepare"}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut)) +
  geom_violin(color="black") +
  scale_fill_manual(values=col_palette) +
  geom_jitter(height=0)

p
```

This is not ideal here because we have many points in each class. Instead, we can reverse the layers.

```{r violinvsjitter_2, exercise=TRUE, exercise.setup="violinprepare"}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut)) + 
  geom_jitter(height=0) +
  geom_violin(color="black") +
  scale_fill_manual(values=col_palette)

p
```

<div class="exo">

- Complete the previous code so that the points have the same color as the violins.

```{r violinjittercol, exercise=TRUE , exercise.setup="violinprepare"}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut, ___)) +
    geom_jitter(height=0) +
    geom_violin(color="black") +
    scale_fill_manual(values=col_palette) +
    ___
p
```

```{r violinjittercol-solution}
col_palette <- c("Ideal" = "#A22200", 
                 "Premium" = "#0871A4", 
                 "Very Good" = "#00B850", 
                 "Good" = "#226666", 
                 "Fair" = "#FF8900")

p <- ggplot(data=diamonds, aes(x=cut, y=price, fill=cut, color=cut))  +
     geom_jitter(height=0) +
     geom_violin(color="black") +
     scale_fill_manual(values=col_palette) +
     scale_color_manual(values=col_palette)

p
```

```{r violinjittercol-check}

gradethis::grade_result_strict(
  pass_if(~ p$labels$x == "cut"),
  pass_if(~ p$labels$y == "price"),
  pass_if(~ p$labels$fill == "cut"),
  pass_if(~ all(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$colour)) == sort(unique(ggplot2::ggplot_build(p)$data[[1]]$fill)))),
  pass_if(~ class(p$layers[[2]]$geom)[1] == "GeomViolin" ), 
  pass_if(~ class(p$layers[[1]]$geom)[1] == "GeomPoint" ) 
)
```

</div>


### End of the section

Thank you for following this tutorial.