---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "Conditional statements"
output:
  learnr::tutorial:
    includes:
      before_body: !expr system.file(file.path("tutorials", "style.html"),package="rtrainer")
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 


```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```

When developing **computer code**, we often need to **ask the system questions** about the **contents of a variable** or multiple variables (*e.g*., has the user submitted an existing file, provided a valid response, or entered a password?). This is the **purpose of conditional structures**.

A conditional structure **allows testing whether a condition is true or false**. These conditional structures generally **evaluate** a test that returns a **logical variable** (true/false):

  - a > 4
  - b < 4
  - 5 < b & b < 10
  - a == 'hamster'
  - a %in% c('hamster', 'dog', 'cat')
  - ...

In the **classic if/else structure, a set of instructions is executed if the condition is true, and a different set of instructions is executed if the condition is false**.

```
b <- 3

if(b < 4) {

  cat("It's true.\n")
  cat("I confirm.")

} else {

  cat("It's false.\n")
  cat("I confirm that it's false!")

}
```

You may also encounter **if without else**.

```

if(b < 4) {
  
  cat("It's true.\n")
  cat("I confirm.")
  
}
```

Or even **if/else if/else**.

```
x <- 3

if(x == 1) {
  
  cat("x is 1.\n")
  
} else if(x == 2) {
  
  cat("x is 2.\n")
  
} else {
  
  cat("x is neither 1 nor 2.\n")
}
```

## The ifelse() Function

In addition to the classic constructions we just reviewed, which are found in many programming languages, R offers a vectorized version. The `ifelse()` function allows iteration over a vector. Its syntax is as follows:

```
ifelse(boolean_vector, value_returned_if_position_is_true,
       value_returned_if_position_is_false)
```

A more concret example:

```{r ifelsevector, exercise=TRUE}
set.seed(123)
x <- sample(c(1,2), size=10, replace=TRUE)
print(x)
x <- ifelse(x == 1, 'chat', 'hamster')
print(x) 
```

<div class="exo">

- You could achieve the same result as the previous one by converting x into a factor, modifying the levels, and then converting x into a character string. What would the code look like?

```{r exoifelse, exercise=TRUE }
set.seed(123)
x <- sample(c(1,2), size=10, replace=TRUE)
```

```{r exoifelse-solution}
x <- factor(x)
levels(x) <- c('cat', 'hamster')
x <- as.character(x)
# or in a single line...
x <- as.character(factor(x, levels = c('cat', 'hamster')))
```
</div>

## Exercises

<div class="exo">

- Download the following tab-delimited file containing information on 10,000 human transcripts. This information was obtained from the [ensembl](https://www.ensembl.org/index.html) database.
- Display the column names.
- Using the `ifelse()` function, create a new column named *mono_exonic* with the value 'mono' if the transcript contains one exon (column `nb_exons`), or 'multi' otherwise.
- Use `table()` to count the occurrences of *mono* and *multi* in the *mono_exonic* column. Store the result in `count_mono_exonic`.
- Represent `count_mono_exonic` with the `barplot()` function.


```{r txGRCh38, exercise=TRUE, exercise.lines=6}
## URL directing to the dataset
url <- "https://zenodo.org/record/8171242/files/Homo_sapiens.GRCh38.110.chr_10k.tsv"
## Adjust timeout settings to prevent interruptions during download  
options(timeout = 1000)
tx_info <- read.table(file=url, header=TRUE, sep="\t", row.name=11)
```

```{r txGRCh38-solution}
## URL directing to the dataset
url <- "https://zenodo.org/record/8171242/files/Homo_sapiens.GRCh38.110.chr_10k.tsv"
## Adjust timeout settings to prevent interruptions during download  
options(timeout = 1000)
tx_info <- read.table(file=url, header=TRUE, sep="\t", row.name=11)
tx_info$mono_exonic <- ifelse(tx_info$nb_exons == 1, 'mono', 'multi')
count_mono_exonic <- table(tx_info$mono_exonic)
barplot(count_mono_exonic)
```

```{r txGRCh38-check }
gradethis::grade_result(
  pass_if(~ table(tx_info$mono_exonic)[1] == 1203)
)
```

</div>


### End of the section

Thank you for following this tutorial.
