---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "La librairie ggplot2 (session 3)"
output:
  learnr::tutorial:
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 

```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```


## Simple Bar Charts

The **geom_bar()** function in ggplot2 is used to create **bar charts**, which are particularly suited for representing categorical data. This function allows you to construct **vertical or horizontal bars** based on the **variables specified in the x and y aesthetics** of the `aes()` function.

In this example, we will create a **bar chart** to represent the frequency of different diamond cuts (`cut`) in the `diamonds` dataset. **The function will count the number of occurrences of each cut: 'Fair', 'Good', 'Very Good', etc.** 
The `position = "dodge"` argument places the bars side by side, making it easier to compare frequencies across different categories. Note below the use of `labs()`, which allows control over the **axis labels**.

```{r convert, exercise=TRUE}
## Load the ggplot2 library
library(ggplot2)
## Next, load the chickwts dataset
data(diamonds, package = "ggplot2")

p <- ggplot(data = diamonds, aes(x = cut, fill = cut)) +
     geom_bar(position = "dodge") +
     labs(title = "Number of Diamond Cuts",
          x = "Cut",
          y = "Count")

print(p)
```

<div class="exo">

- Modify the following code to associate a unique color with each bar.

```{r barplotcol, exercise=TRUE , exercise.setup="convert"}
p <- ggplot(data = diamonds, aes(x = cut)) +
        geom_bar(position = "dodge") +
       labs(title = "Nombre de type de clarté par Coupe",
            x = "Coupe",
            y = "Nombre")
print(p)
```

```{r barplotcol-solution}
col_palette <- c("Ideal" = "#A22200", 
                  "Premium" = "#0871A4", 
                  "Very Good" = "#00B850", 
                  "Good" = "#226666", 
                  "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = cut, fill=cut)) +
            geom_bar(position = "dodge") +
            labs(title = "Nombre de type de clarté par Coupe",
                     x = "Coupe",
                     y = "Nombre",
                     fill="Coupe") +
            scale_fill_manual(values=col_palette)
print(p)
```

```{r barplotcol-check}
gradethis::grade_result_strict(
  pass_if(~ length(unique(ggplot2::ggplot_build(p)$data[[1]]$fill)) > 1),
  pass_if(~ "GeomBar" %in% class(p$layers[[1]]$geom))  
)
```

</div>

## Exemple à deux variables catégorielles

Dans cet exemple plus complexe, nous allons créer un graphique à barres en utilisant `cut` comme axe x et `clarity` comme axe y pour explorer la fréquence des différentes clartés de diamants (*clarity*) pour chaque type de coupes (`cut`). Le principe est de dire que les barres seront colorés (`fill`) en fonction de la variable `clarity`.

Nous utiliserons l'argument `position = "stack"` (*i.e.* empilée) pour placer les comptes des différentes clartés les une sur les autres.

```{r complexebarplot, exercise=TRUE, exercise.setup="convert" }
p <- ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "stack") +
       labs(title = "Nombre de type de clarté par Coupe",
            x = "Coupe",
            y = "Nombre")

print(p)
```

Note that `clarity` is an **ordinal categorical variable**, and ggplot naturally chooses a **discrete color gradient** to represent it.

```{r clarity, exercise=TRUE, exercise.setup="convert" }
head(diamonds$clarity)
```

As in the previous examples, you can change the colors with `scale_fill_manual`. However, at this stage, we can introduce a new function, `scale_fill_brewer()`, which requires installing the `RColorBrewer` library. For this function, you need to choose the name of one of the palettes displayed by `RColorBrewer::display.brewer.all()`. You pass the palette name to the function (*e.g.*, `scale_fill_brewer(palette='Purples')`).

```{r RColorBrewer, exercise=TRUE, fig.height=7, exercise.setup="convert"}
if(!require('RColorBrewer')){install.packages('RColorBrewer')}
RColorBrewer::display.brewer.all()
```

There are **three types of palettes in `RColorBrewer`: sequential, diverging, and qualitative**.

- **Sequential palettes** are suitable for ordered data that progresses from the lowest to the highest value.
- **Diverging palettes** emphasize critical middle values and extremes at both ends of the data range.
- **Qualitative palettes** are best suited for representing nominal or categorical data.


<div class="exo">

- Test different sequential palettes from RColorBrewer (*e.g.* Blues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu, PuBuGn, PuRd, Purples, RdPu, Reds, YlGn, YlGnBu, YlOrBr, YlOrRd) for the representation.

```{r scalegradient, exercise=TRUE, exercise.setup="RColorBrewer"}
p <- ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "stack", color="black") +
       labs(title = "Nombre de type de clarté par Coupe",
            x = "Coupe",
            y = "Nombre") + scale_fill_brewer(palette='___')
print(p)
```

```{r scalegradient-solution}
p <- ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "stack", color="black") +
       labs(title = "Nombre de type de clarté par Coupe",
            x = "Coupe",
            y = "Nombre") + scale_fill_brewer(palette='Purples')
print(p)
```

</div>

## Bar Positioning

In ggplot2, the **'position' argument** of `geom_bar` **determines the positioning of the bars** and can drastically change the perception:

- **'Stack' stacks the bars** to represent cumulative values, useful for showing totals and proportions of subcategories.
- **'Dodge' places the bars side by side without overlap**, ideal for directly comparing values between categories.
- **'Dodge2' is similar to 'dodge' but further separates the bars** based on another variable, creating side-by-side groups.
- **'Fill'** fills the entire bar. Useful for visualizing proportions/percentages.
- **'Identity'** places the bars in front of each other. Be cautious, as this is rarely what you want.
- **'Jitter'** adds a bit of noise to the x-axis. In the context of bar charts, this argument has little value. Again, the bars are placed in front of each other.

<div class="exo">

- In the diagram below, change the position successively to 'stack', 'dodge', and 'fill'.

```{r scalegradient2, exercise=TRUE , exercise.setup="RColorBrewer"}

p <- ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "___", color="black") +
       labs(title = "Number of Clarity Types by Cut",
            x = "Coupe",
            y = "Nombre") + scale_fill_brewer(palette='Oranges')
print(p)
```

```{r scalegradient2-solution}
ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "stack", color="black") +
       labs(title = "Number of Clarity Types by Cut",
            x = "Coupe",
            y = "Nombre") + scale_fill_brewer(palette='Purples')

ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "dodge", color="black") +
       labs(title = "Number of Clarity Types by Cut",
            x = "Coupe",
            y = "Nombre") + scale_fill_brewer(palette='Purples')

ggplot(data = diamonds, aes(x = cut, fill=clarity)) +
        geom_bar(position = "fill", color="black") +
       labs(title = "Number of Clarity Types by Cut",
            x = "Coupe",
            y = "Proportion") + scale_fill_brewer(palette='Purples')

```

</div>

## Consolidation Exercises

### Iris Dataset

<div class="exo">

- Create a plot like the one shown below that compares the sepal lengths (`Sepal.Length`) for each flower species (`Species`) in the iris dataset (`data(iris)`). The plot, stored in a variable `p`, should display the distributions as boxes (`geom_boxplot()`) with jittered points (`geom_jitter()`) to visualize each individual observation. The axes should be renamed "Species" and "Sepal Lengths." Each species should be associated with a unique box color ('fill'). 


```{r, echo=FALSE, fig.height=4}
library(ggplot2)
data(iris)
# Example of comparing distributions with violins
gg <- ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter() +
  scale_fill_brewer(palette='Dark2') +
  labs(title = "Comparison of Sepal Lengths by Species",
       x = "Species",
       y = "Sepal Lengths",
       fill = "Species") 

gg
```

```{r iris, exercise=TRUE , exercise.setup="RColorBrewer"}
data(iris)
```

```{r iris-solution}
data(iris)
# Example of comparing distributions with violins
p <- ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter() +
  scale_fill_brewer(palette='Dark2') +
  labs(title = "Comparison of Sepal Lengths by Species",
       x = "Species",
       y = "Sepal Lengths",
       fill = "Species") 

print(p)
```

```{r iris-check}

gradethis::grade_result_strict(
  pass_if(~ p$labels$x == "Species"),
  pass_if(~ p$labels$y == "Sepal Lengths"),
  pass_if(~ length(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$fill))) == 3),
  pass_if(~ class(p$layers[[1]]$geom)[1] == "GeomBoxplot"),
  pass_if(~ class(p$layers[[2]]$geom)[1] == "GeomPoint")
)
```
</div>

### ToothGrowth Dataset

<div class="exo">

- Create a violin plot like the one shown below to compare the distribution of tooth lengths (`len`) based on the method of vitamin C administration (`supp`) in the `ToothGrowth` dataset.

- The categories 'OJ' and 'VC' should be replaced with 'Orange juice' and 'Ascorbic acid'.
- Add "rugs" (small lines) to the plot (`geom_rug()`) to visualize the data distribution along the y-axis.
- The axes should be renamed "Supplements" and "Tooth Lengths". Each violin should have its own unique color ('fill').


```{r, echo=FALSE, fig.height=4}
data(ToothGrowth)
levels(ToothGrowth$supp) <- c('Orange juice', 'Ascorbic acid')
gg <- ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp, color=supp)) +
  geom_violin(alpha = 0.7, color="black") +
  geom_rug() +
  scale_fill_brewer(palette='Accent') +
  scale_color_brewer(palette='Accent') +
  labs(title = "Comparison of Tooth Length by Dose",
       x = "Supplements",
       y = "Tooth Lengths",
       fill = "Supplements",
       color = "Supplements") 
gg
```

```{r ToothGrowth, exercise=TRUE , exercise.setup="RColorBrewer"}
data(ToothGrowth)
```

```{r ToothGrowth-solution}
data(ToothGrowth)
levels(ToothGrowth$supp) <- c('Orange juice', 'Ascorbic acid')
p <- ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp, color=supp)) +
  geom_violin(alpha = 0.7, color="black") +
  geom_rug() +
  scale_fill_brewer(palette='Accent') +
  scale_color_brewer(palette='Accent') +
  labs(title = "Comparison of Tooth Length by Dose",
       x = "Supplements",
       y = "Tooth Lengths",
       fill = "Supplements",
       color = "Supplements") 
print(p)
```

```{r ToothGrowth-check}

gradethis::grade_result_strict(
  pass_if(~ p$labels$x == "Supplements"),
  pass_if(~ p$labels$y == "Tooth Lengths"),
  pass_if(~ length(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$fill))) == 2),
  pass_if(~ class(p$layers[[1]]$geom)[1] == "GeomViolin"),
  pass_if(~ class(p$layers[[2]]$geom)[1] == "GeomRug"),
  pass_if(~ "Orange juice" %in%  as.matrix(p$data)),
  pass_if(~ 'Ascorbic acid'  %in%  as.matrix(p$data))
)
```
</div>

### Palmerpenguins Dataset

For this example, we will use the `penguins` dataset from the `palmerpenguins` package. This dataset contains information about penguins from the Palmer Archipelago in Antarctica.

<div class="exo">

- Using the penguins dataset, generate a plot `p` identical to the diagram below. In this example, we use `flipper_length_mm` (flipper length) on the x-axis, `bill_length_mm` (bill length) on the y-axis, and `body_mass_g` (body mass) for the size of the points, with `species` (species) for the color of the points. You must also ensure the axis labels for the x and y axes match the given names.

```{r, echo=FALSE, fig.height=4, }
library(palmerpenguins)
data("penguins")
penguins <- na.omit(penguins)

# Create a custom color palette
palette_couleurs <- c("dodgerblue", "darkorange", "forestgreen")

# Create the colored bubble plot
p <- ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm, size = body_mass_g, color = species)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = palette_couleurs) +
    labs(x = "Flipper Length (mm)", 
         y = "Longueur du bec (mm)", 
         color = "Espèce", 
         size = "Masse corporelle (g)") 
print(p)

``` 

```{r penguins, exercise=TRUE }
library(palmerpenguins)
data("penguins")
penguins <- na.omit(penguins)

```

```{r penguins-solution}
library(palmerpenguins)
data("penguins")
penguins <- na.omit(penguins)

# Create a custom color palette
palette_couleurs <- c("dodgerblue", "darkorange", "forestgreen")

# Create the colored bubble plot
p <- ggplot(penguins, aes(x = flipper_length_mm, y = bill_length_mm, size = body_mass_g, color = species)) +
    geom_point(alpha = 0.7) +
    scale_color_manual(values = palette_couleurs) +
    labs(x = "Flipper Length (mm)", 
         y = "Bill Length (mm)", 
         color = "Species", 
         size = "Body Mass (g)") 
print(p)
```

```{r penguins-check }

gradethis::grade_result_strict(
  pass_if(~ p$labels$x == "Flipper Length (mm)"),
  pass_if(~ p$labels$y == "Bill Length (mm)"),
  pass_if(~ length(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$colour))) == 3),
  pass_if(~ class(p$layers[[1]]$geom)[1] == "GeomPoint"),
  pass_if(~ sum(ggplot2::ggplot_build(p)$data[[1]]$size) > 1300)
)
```
</div>

### End of the section

Thank you for following this tutorial.