---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "For loops"
output:
  learnr::tutorial:
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 


```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(rtrainer)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```

## Principle and Syntax

### Introduction

When developing **computer code**, we often need to **iterate over elements stored in objects** that are themselves stored in memory (*e.g.*, vectors, matrices...). Depending on the problem, we might want to iterate over users, prices, products, chromosomes, genes... We have already seen that:

- **apply** family functions are a solution for applying functions to the elements of an object.

- Similarly, **vectorization** allows for implicitly iterating over objects to perform operations (e.g., `x + 1` to iterate over all elements of a vector $x$ and add `1` to them).

R also offers the **classic for loop structure** found in many programming languages. The syntax of the for loop is given below.


```
for(element in my_object){ 
    instruction 1
    instruction 2
    ...
    instruction n
    }
```

  - `element` is a variable whose **name is freely chosen** but should be explicit. The object `element` will be **instantiated (defined in memory) when the loop starts**. Therefore, it does not need to be declared/created beforehand.
  - `my_object` is a **pre-existing object**. It must have been declared earlier in the code.
  - `my_object` must be **an iterable object**. It should be designed to be iterated over. This includes vectors, matrices, data.frames, lists, etc.


## Exercices

```{r createfile, echo=FALSE, eval=TRUE, include=FALSE }
tmp_path <- rtrainer::write_fasta_fragilaria()
```

<div class="exo">

- Given the following path (`tmp_path`), list (`dir()`) all the files in this folder. Check the content of one of them with `file.show()`.
- Using a **for loop**, read them with `scan()` (`what="character"`, `sep = "\n"`) and iteratively store their content (*i.e.*, the object returned by `scan()`) in a list named `my_seq`.

```{r createfile_exo, exercise=TRUE , exercise.setup="createfile"}
tmp_path
```

```{r createfile_exo-solution}
my_seq <- list()
for(i in file.path(tmp_path, dir(tmp_path))){
  my_seq[[basename(i)]] <- scan(file=i, what="character", sep = "\n")
}
```

```{r createfile_exo-check }
gradethis::grade_result(
  pass_if(~ length(my_seq) > 10),
  pass_if(~ "AGTGACCGTTACGAATCTGGTGTAATTCCTTACGCTAAAATGGGTTACTGGGATGCTTCATACGCTGTAAAAACTACTGA" %in% unlist(lapply(my_seq, "[", 2)))
)
```
</div>


### End of the section

Thank you for following this tutorial.
