---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "Conversions"
author: "D. Puthier"
output:
  learnr::tutorial:
    includes:
      before_body: !expr system.file(file.path("tutorials", "style.html"),package="rtrainer")
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 


```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```

## Conversion Functions

### Principle and Dedicated Functions

In some cases, an object can be converted into another type. These conversions (also known as coercion or *cast*) allow you to **change the format** of an object to pass it to a function that would otherwise **not be compatible with its original format**. The functions used for this purpose generally have the prefix "as.". As you can see from the following command, there are many such functions.

Here are some examples:

- `as.factor()` : to convert to a factor.
- `as.data.frame()` : to convert to a data.frame.
- `as.matrix()` : to convert to a matrix.
- `as.list()` : to convert to a list.
- `as.numeric()` : to convert to a numeric vector.
- `as.character()` : to convert to a character vector.

```{r convertion, exercise=TRUE}
apropos("^as\\.")
```

Additionally, there are conversion functions that do not start with the prefix "as.", such as `unlist()`, which transforms a list into a vector.

```{r unlist, exercise=TRUE}
my_list <- list(A=1:5, B=letters[1:5])
unlist(my_list)
```

### Convert to Modify a Result

This exercise aims to demonstrate one of the key advantages of conversion. This is one of R's strengths. Some **functions are polymorphic**, meaning their behavior **depends on the type of argument provided to the function**.

<div class="exo">

- Given the following matrix. Use the `boxplot()` function on the matrix and on the matrix converted to a vector (`as.vector()`).

```{r boxplotas, exercise=TRUE}
set.seed(456)
mat <- matrix(c(rnorm(n=20, mean=5, sd=2),
                rnorm(n=20, mean=3, sd=1),
                rnorm(n=20, mean=2, sd=2)),
                ncol=3, byrow = FALSE)
         
```

```{r boxplotas-solution}
boxplot(mat)
boxplot(as.vector(mat))
```
</div>

```{r quizz_1, echo=FALSE }
learnr::question("What do you observe",
  answer("The function behaves differently depending on the argument passed to it.", correct = TRUE),
  answer("The `boxplot()` function behaves differently depending on the argument it returns."),
  answer("If you pass a vector to the `boxplot()` function, it produces multiple boxplots."),
  answer("If you pass a vector to the `boxplot()` function, it produces a single boxplot.", correct = TRUE),
  answer("If you pass a matrix to the `boxplot()` function, it produces multiple boxplots.", correct = TRUE),
  allow_retry = TRUE 
)
```

```{r quizz_2, echo=FALSE }
learnr::question("What is the benefit of this behavior?",
  answer("Avoiding the need to multiply functions.", correct = TRUE),
  answer("There is none."),
  answer("Easily changing the result by modifying the type of the input object.", correct = TRUE),
  answer("Making the code more complicated."),
  answer("Making the language easier to learn.", correct = TRUE),
  allow_retry = TRUE 
)
```

<div class="exo">

- Given the same matrix, apply the `plot()` function on:

  - the matrix $mat$ passed as the sole argument.
  - the first column of $mat$ passed as the sole argument (drop=TRUE).
  - the first column of $mat$ passed as the first argument and the second column of $mat$ passed as the second argument.

```{r plotas_2, exercise=TRUE, exercise.setup="boxplotas"}

```

```{r plotas_2-solution}
plot(mat)
plot(mat[, 1])
plot(mat[, 1], mat[, 2])
```
</div>


```{r quizz_3, echo=FALSE }
learnr::question("What is R's behavior when using the `plot()` function?",
  answer("If the first argument is a matrix, R randomly represents two columns of the matrix?"),
  answer("If the inputs are two vectors a and b, R displays for each position $i$ $a_{i}$ *versus* $b_{i}$.", correct = TRUE),
  answer("If the first argument is a vector, R randomly displays the values of the vector."),
  answer("If the first argument is a matrix, R represents column 1 on the x-axis *versus* column 2 on the y-axis.", correct = TRUE),
  answer("If the first argument is a vector (and the y argument is not provided), R displays the position in the vector on the x-axis and the corresponding value on the y-axis.", correct = TRUE),
  answer("If the first argument is a matrix, R displays the position in the vector on the x-axis and the corresponding column on the y-axis."),
  allow_retry = TRUE 
)
```


### End of the section

Thank you for following this tutorial.

