---
# theme
# “bootstrap”, “cerulean”, “cosmo”, “darkly”, “flatly”, “journal”, “lumen”, “paper”, 
# “readable”, “sandstone”, “simplex”, “spacelab”, “united”, “yeti”
# highlight: `default`, `tango`, `pygments`, `kate`, `monochrome`, `espresso`, `zenburn`, 
# `haddock`, `breezedark`, `textmate`, `arrow`, or `rstudio` or a file with extension `.theme`.

title: "The ggplot2 library (session 1)"
output:
  learnr::tutorial:
    theme: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 

```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(png)
library(ggplot2)
library(RCurl)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```


## The ggplot2 Library

### Limitations of Basic Graphics in R   

R provides **numerous graphical functions**, some of which we have encountered: `hist()`, `plot()`, and `barplot()`. However, **these basic graphical functions** suffer from several limitations: 

- Their syntax/arguments are not standardized.
- Modifying certain graphical parameters can be complex and unintuitive.
- Although it is possible to overlay plots (e.g., `points()`, `abline()` on a diagram created with `plot()`), the number of combinations is limited.
- Moreover, when performing descriptive statistics, we often want to **partition the graphical window** (*i.e.*, create **facets**) based on the levels of a categorical (*i.e.*, qualitative) or ordinal variable. Creating such graphics quickly becomes extremely complicated with the basic tools.

### The ggplot Library

To rethink the design of graphics in R, **Hadley Wickham** developed the **ggplot2** library, which has quickly become **popular and essential** in the world of bioinformatics and data analysis. This library is now considered **one of R's true strengths** compared to languages like **Python** or **Julia**.

One of the unique aspects of the ggplot2 library is that its development is based on a model proposed by **Leland Wilkinson** in his book **The Grammar of Graphics**. This book offers insights into the components of graphics. According to **Leland Wilkinson**:

- A graphic represents **a correspondence** between **data** and **aesthetic attributes (color, shape, size...) of geometric objects (points, lines, bars...)**.
- The graphic can also include **data transformations** (*e.g.*, logarithmic scale) and is drawn in a specific **coordinate system** (*e.g.*, Cartesian or polar).
- **Facets** can be used to generate the same graphic for **different subsets of the dataset**.

It is the combination of these independent components that forms a graphic. In ggplot2, these different components are modeled **as layers** that can be **added** to one another using **the addition operator**. The success of ggplot2 has led many developers to produce **extensions**, creating an **extremely rich suite of graphical functions** today.


```{r, echo=FALSE, fig.height=6, fig.width=6}
library(png)
pp <- readPNG(getURLContent("https://zenodo.org/records/15083675/files/ggbasics.png"))
plot.new() 
rasterImage(pp,0,0,1,1)
```

You can find a ggplot2 cheatsheet as a complement to this course at [this address](https://thinkr.fr/pdf/ggplot2-french-cheatsheet.pdf).

### The Dataset

To start, we will use the "diamonds" dataset from the **ggplot2** library. The **diamonds** dataset contains various information about 54,000 diamonds. It is part of the `ggplot2` package. The following details are available:

- Carat, size, color, clarity, depth, table, price, x, y, z

```{r, echo=FALSE, fig.height=5}
library(png)
library(RCurl)
pp <- readPNG(getURLContent("https://zenodo.org/records/15083675/files/diamonds.png"))
plot.new() 
rasterImage(pp,0,0,1,1)
```

Let's load the dataset.

```{r dataset, exercise=TRUE }
## Load the ggplot2 library
library(ggplot2)
## Next, load the chickwts dataset
data(diamonds)
```

The first lines of this dataset are the following :

```{r head, exercise=TRUE, exercise.setup="dataset"}
head(diamonds)
```

The types of the columns are the following.

```{r type, exercise=TRUE, exercise.setup="dataset" }
str(diamonds)
```

Note that the `diamonds` object could be a `data.frame`. In fact, it is a `tibble`, which is very close to a `data.frame`.

```{r convert, exercise=TRUE, exercise.setup="dataset" }
class(diamonds)
```

## Scatterplots

There are more than 50 types of plots (geometries) available in `ggplot2`, and even more if we consider ggplot2 extensions.

We will start with a simple scatterplot. To create our scatterplot, we will use the `geom_points()` geometry. This requires:

- *(i)* **Data** in `data.frame` or `tibble` format.
- *(ii)* **Associating this data** with a ggplot object (`ggplot(data=diamonds)`).
- *(iii)* Defining **the columns of the dataframe to use and associate them with the aesthetics of a selected geometrical objects**.

For a scatterplot (`geom_points()`), two aesthetics are mandatory: $x$ (x-axis) and $y$ (y-axis). The $x$ and $y$ variables are typically numeric. Here, we will associate the `carat` and `price` columns, both of which are numeric, to the `x`and `y`aesthetic using the `aes()` function.

Then we apply the geometry using `+ geom_points()` which adds a layer to the diagram.

```{r firstgg, exercise=TRUE , exercise.setup="convert" }
# We create the plot by passing 
# the data.frame (diamonds) and the result 
# of the aes() function to ggplot. 
# The aes() function associates column names 
# with aesthetic elements.
p <- ggplot(data=diamonds, aes(x=carat, y=price)) + geom_point()
p
```

You can modify the aesthetics of the geometric shape (here, points) by passing arguments to the geometry function (here `geom_points()`).

For the **geom_points** geometry, the following **aesthetics** are accepted:

| **Aesthetic**  |                                   **Purpose**                                   |              **Value**              |
|:--------------:|:-------------------------------------------------------------------------------:|:-----------------------------------:|
|       x        |                                Values for the x-axis                           |         Numeric or factor           |
|       y        |                                Values for the y-axis                           |         Numeric or factor           |
|     alpha      |                                Opacity of points                               |         Numeric between 0 and 1     |
| colour or color| Color of points or borders (if points have borders)                            |         A color                     |
|      fill      | Filling, if the point has a fill                                               |         A color                     |
|      shape     | Shapes of points                                                              |         An integer                  |
|      size      | Sizes of points                                                               |         Numeric                     |
|     stroke     | Border sizes                                                                  |         Numeric                     |
|      ...       |                                                                               |                                     |

For example, you can change the color and size of the points as follows:


```{r gg2, exercise=TRUE , exercise.setup="convert" }
p <- ggplot(data=diamonds, aes(x=carat, y=price)) + 
       geom_point(size=2, colour="red")
p
```

## Point Shapes (Aesthetic *shape*)

The correspondences between point shapes and the numeric variables passed are shown below:


```{r, echo=FALSE, fig.height=5}
library("png")
library(RCurl)
pp <- readPNG(getURLContent("https://zenodo.org/records/15083675/files/point.png"))
plot.new() 
rasterImage(pp,0,0,1,1)
```

Other point shapes are possible, for example, by using the [ggstar](https://cran.r-project.org/web/packages/ggstar/vignettes/ggstar.html) library:

```{r, echo=FALSE, fig.height=5}
library(png)
pp <- readPNG(getURLContent("https://zenodo.org/records/15083675/files/ggstar.png"))
plot.new() 
rasterImage(pp,0,0,1,1)
```

More generally, any image can be used as a point (using [ggimage](https://yulab-smu.top/pkgdocs/ggimage.html)).

<div class="exo">

### Exercise

- Using point shape 22 (`shape=22`), create a scatterplot (`geom_point()`) and color the interior and borders of the points with colors of your choice. Adjust the opacity to better visualize overlapping points.

```{r colorgg2, exercise=TRUE , exercise.setup="convert"}
colors()
p <- ggplot(data=diamonds, aes(x=carat, y=price))
```

```{r colorgg2-solution}
p <- ggplot(data=diamonds, aes(x=carat, y=price))
p + geom_point(size=2, shape=22, fill="violet", color="darkviolet", alpha=0.3)
```


</div>


## Point Colors (Aesthetics *colour* and *fill*)

**Colors** play an **essential role in data visualization**, and R provides a **wide range of options to manipulate and use colors** in your plots and analyses.

R has a convenient function called `colors()` that displays the names of the colors available in R. This function returns a vector containing the predefined color names. Here's an example to display some of these colors:


```{r colors, exercise=TRUE}
head(colors(), 20)
```

In addition to color names, it is common to represent colors using the **hexadecimal format**. In this format, each color is represented by a combination of **six hexadecimal digits**, corresponding to the **amount of red (R), green (G), and blue (B) in the color**. Each color has an intensity level between 00 (minimum) and FF (maximum). For example, green is represented by #00FF00, red by #FF0000, black by #000000, and white by #FFFFFF.

In R, you can specify colors in hexadecimal format by simply using the # prefix followed by the appropriate hexadecimal combination. For example:

```{r hexa, exercise=TRUE, exercise.setup="convert"}
p <- ggplot(data=diamonds, aes(x=carat, y=price)) + 
      geom_point(size=2, shape=22, fill="#7F00FF", color="#9400D3", alpha=0.3)
p
```

## All 'Aesthetics' can be linked to variables

The control of aesthetics in ggplot2 provides remarkable flexibility by allowing you to customize plots with **fixed values** as we have seen (*e.g.*, color, size, opacity) but also with **variables**. This unique feature enables dynamic and customized graphics based on your data. You can modify the color, size, shape (...) of points based on the values of a column. To do this, you **associate additional variables with aesthetics using the `aes()` function**.

For example, we can use a variable like **`cut` to control the color of the points** and another variable **like `depth` to adjust their size**:

```{r varasesthetics, exercise=TRUE, exercise.setup="convert"}
p <- ggplot(data = diamonds, aes(x = carat, 
                                 y = price, 
                                 color = cut, 
                                 size = depth)) +
         geom_point(alpha=0.5)

p
```

If you want to customize the colors, you can add a layer to the plot using the functions `scale_color_manual()` or `scale_fill_manual()` to manually define the colors for the `color` and `fill` aesthetics.

<div class="exo">

- Complete the following code to associate custom colors with the plot $p$:

```{r colpoints, exercise=TRUE , exercise.setup="convert"}
col_pal <- c("Ideal" = "#A22200", 
             "Premium" = "#0871A4", 
             "Very Good" = "#00B850", 
             "Good" = "#226666", 
             "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = carat, 
                                 y = price, 
                                 color = cut, 
                                 size = depth)) +
  geom_point(alpha=1, shape=16) +
  scale_color_manual(values = ___)

p
```

```{r colpoints-solution}
col_pal <- c("Ideal" = "#A22200", 
             "Premium" = "#0871A4", 
             "Very Good" = "#00B850", 
             "Good" = "#226666", 
             "Fair" = "#FF8900")

p <- ggplot(data = diamonds, aes(x = carat, 
                                 y = price, 
                                 color = cut, 
                                 size = depth)) +
  geom_point(alpha=1, shape=16) +
 scale_color_manual(values = col_pal)

p
```

```{r colpoints-check}
 
gradethis::grade_result_strict(
  pass_if(~ any(sort(unique(ggplot2::ggplot_build(p)$data[[1]]$colour)) != c("#21908CFF", 
                                                                             "#3B528BFF", 
                                                                             "#440154FF", 
                                                                             "#5DC863FF", "#FDE725FF")))
)
```

</div>


ggplot2 offers **great flexibility** in **controlling the aesthetics of your graphs**. You can use **fixed values** to create static graphs, but using **variables allows even more dynamic, data-driven customization**. 

### End of the section

Thank you for following this tutorial.