---
title: "Preambles"
output:
  learnr::tutorial:
    includes:
      before_body: !expr system.file(file.path("tutorials", "style.html"),package="rtrainer")
    theme: default
    highlight: default
    fig_caption: yes
    self_contained: true
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 4
    number_sections: false
    progressive: true
  html_document:
    theme: cosmo
    fig_caption: yes
    self_contained: yes
    toc: yes
    toc_float: 
      toc_collapsed: false
    toc_depth: 3
    number_section: true
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_md: no
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  slidy_presentation:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    highlight: tango
    incremental: no
    keep_md: no
    self_contained: yes
    slide_level: 2
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
runtime: shiny_prerendered
---

<!--  
Here the parameters about the documents.
https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf 
--> 


```{css, echo=FALSE}

```

<script language="JavaScript" type="text/javascript">
          
          function sizeTbl2(h,i) {
          var tbl = document.getElementById("section-" + i);
          tbl.style.display = h;
          }

</script>



<style>
.exo {
  border-radius: 5px;
  margin-top: 5px;
  margin-bottom: 5px;
  padding-top: 5px;
  padding-bottom: 5px;
  padding-left: 5px;
  padding-right: 5px;
  background-color: #fcede3;
  color: rgb(51, 51,153);
}
.tips {
       padding-top: 5px;
       padding-bottom: 5px;
       padding-left: 5px;
       padding-right: 5px;
       border: 1px dashed #2f6fab;
       background-color: #EEFFEE;
}
.solution {
            margin-top: 5px;
            margin-bottom: 5px;
            padding-top: 5px;
            padding-bottom: 5px;
            padding-left: 5px;
            padding-right: 5px;
            border: 1px dashed #FFFFFF;
            background-color: #EEEEFF;
            color: #0000BB;
            font-size: 11px;
            hidden=true;
}
</style>


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, comment="")
gradethis::gradethis_setup()
```


## A few words about R

The rise of **high-throughput biological analysis techniques** has led to a deluge of data, requiring biologists to handle datasets with thousands to billions of values. Flexible software for **data collection, processing, and statistical analysis** is now essential.

R, developed by **Robert Gentleman and Ross Ihaka**, is a free statistical analysis tool available on Unix, Linux, Windows, and Mac. Its **open-source** nature allows continuous evolution through contributions from an active **developer community**.

Widely used by scientists in fields like biology, statistics, and geography, R promotes collaboration and the exchange of tools, driving its success. Its flexibility and interactive command-line execution make it accessible, though new users may need time to master it.

Learning R pays off quickly, thanks to its vast library of packages tailored for diverse tasks, from classical biology to high-throughput methods. Additionally, R offers advanced graphical tools for customizable outputs in formats like .jpg, .png, .pdf, and .html, making data sharing and presentation seamless.

## Object creation

### Assignment operators

R is an object-based language (*vector, matrix, data.frame, list, factor,*...). We generally use the '<-' (arrow) operator to create objects. To begin with, let's create a simple "vector" object (a kind of array) that will contain a numerical value. Here, the code "x <- 15" could read "the variable x takes the value 15" or "we assign 15 to x".

```{r assign0, exercise=TRUE}
x <- 15
print(x)
```

The arrow assignment operator '<-' can be substituted by the '=' operator (this syntax is often used by R language beginners). However, it is **always preferable to use** '<-'. This is the recommendation of Hadley Wickham, developer of numerous function libraries, in his [*tidyverse style guide*](https://style.tidyverse.org/) (*cf* section 2.7).

```{r assign1, exercise=TRUE, exercise.setup = "assign0"}
x = 15
print(x)
```

The following form (*i.e* the inverted arrow) is also accepted. However, it is rarely recommended.

```{r assign0b, exercise=TRUE}
15 -> x
print(x)
```

<div class="exo">
- Write a code to store 3 in a variable x, 5 in y and then the sum (operator *+*) of x and y in z. Print the contents of the 3 variables.

```{r assign2, exercise=TRUE, exercise.setup = "assign1"}

```


```{r assign2-solution, eval=TRUE, echo=FALSE}
x <- 3
y <- 5
z <- x + y
print(x)
print(y)
print(z)
```

```{r assign2-check, eval=TRUE, echo=FALSE}
gradethis::grade_result_strict(
  pass_if(~ x == 3),
  pass_if(~ y == 5),
  pass_if(~ z == 8)
)
```
</div>

### How to name objects

There are several recommendations for naming objects.

- It is generally recommended to use explicit, lower-case variable names. 
- We'll follow the recommendations of [Hadley Wickman](http://adv-r.had.co.nz/Style.html) and use an "underscore" character (_) to separate words in a variable name. 
- Variable names should be in English as all other language elements are in English (e.g function names.  
- Try to keep names concise and meaningful. Here are a few examples.


```{r assign3, exercise=TRUE}
age <- 20
average_size <- 130
nb_iteration <- 100
nb_genes <- 20e3
```

**In this tutorial, we make some exceptions** by using variable names like `x`, `y`, or `z`. This keeps the examples concise and straightforward. However, such naming is not advised for production code.

## Code elements

### Comments

Code may be commented using the "#" character. Any command following this character will not be interpreted. This allows you to annotate your code (that is explain how it works for yourself and others).

```{r assign3b, exercise=TRUE}
# This is a comment
z <- 10 # This is another comment
```


### Separing instructions

Instructions can be separated by a carriage return or by the **';'** character.

```{r assign4, exercise=TRUE}
x <- 12; y <- 13
print(x)
print(y)
```

## Manage objects in memory

### Listing and destroying objects

Created objects are stored in the computer's RAM. At this stage, they are not stored on disk. If you quit the program without saving, they will be destroyed. 

To find out which objects are available (*i.e* have been created at a given time), you can **list** the objects in memory with the **ls()** function. 

```{r assign5, exercise=TRUE}
# Let's create some objects
x <- 1; y <- 2; z <- 3
# List the objects stored in memory
ls()
```

Objects can be destroyed using the **rm()** function (**remove**).

```{r assign5b, exercise=TRUE, exercise.setup="assign5"}
# Destroy x 
rm(x)
ls()
```

### Predefined variables

There are a number of predefined variables  in the R environment. These include `LETTERS` (upper-case alphabet), `letters` (lower-case alphabet), `pi` (...). We'll use them in our code according to our needs.

```{r letters, exercise=TRUE}
letters
```
```{r letters_2, exercise=TRUE}
LETTERS
```
```{r pi, exercise=TRUE}
LETTERS
```


## Functions

### Functions ? 

In R, you can call functions to **query objects and perform actions on them**. Functions take the following form:

```
functionname(arg1= a, arg2= b,...)

# Above arg1 and arg2 (...) correspond to argument names
# a and b correspond to the objects/values you wish to pass to the 
# 'functionname' function to perform a task using them.

```
We've already used the `ls()` and `rm()` functions. Lots of other functions exist. For example, you can use the `round()` function to rounds the values of a vector to the specified number of decimal places. 


```{r function0, exercise=TRUE }
x <- 170.4578
round(x, 2)
```

<div class="exo">

- Use the `round()` function with argument round x to 1 decimal places. Store the result in x (*i.e.* overwrite `x`).

```{r lsexo, exercise=TRUE}
x <- 170.4578
```

```{r lsexo-solution}
x <- 170.4578
x <- round(x, 1)
```


```{r lsexo-check}
gradethis::grade_result_strict(
  pass_if(~ x == 170.5)
)
```
</div>


### Examples of function usage

The `c()` (**combine**) function is used to combine values (e.g. numerical) into a *vector*.  

**NB:** below 1:8 produces a vector containing all integers from 1 to 8. 

```{r function1, exercise=TRUE}
# Integers from 1 to 8
x <- 1:8
# Combine -50, x (i.e. 1 to 8) and 70 
# into a new vector y
y <- c(-50, x, 70)
print(y)
```

The `length()` function lets you find out the size of a vector.

```{r function_1_1, exercise=TRUE, exercise.setup="function1"}
# The length of y (i.e the number of elements in y)
print(length(y))
```

The `is()` and `class()` functions are used to find out the type of an object. Knowing the type of an object is important for knowing which functions can be applied to it. 

```{r function2, exercise=TRUE, exercise.setup="function1"}
is(y)
class(y)
```

<div class="exo">

- Store the type of x (`class()`) in z.

```{r is, exercise=TRUE}
x <- matrix()
```

```{r is-solution}
x <- matrix()
z <- class(x)
```


```{r is-check}
gradethis::grade_result_strict(
  pass_if(~ z[1] == "matrix")
)
```

</div>

### Function arguments

The `round()` function has two arguments: *x* and *digits* (which controls the number of decimal places).


```{r}
args(round)
```

```{r fonction3b, exercise=TRUE}
y <- c(2.567, 3.987, 4, 10.34566)
round(x=y, digits = 2)
```

Note that **argument names can be abbreviated** (note that abbreviation can be ambiguous when there are many arguments with the same prefix).

```{r fonction4, exercise=TRUE}
y <- c(2.567, 3.987, 4, 10.34566)
round(x=y, d = 2)
```

**Argument names can also be omitted**. However, they must be in the expected order.

```{r fonction4b, exercise=TRUE}
y <- c(2.567, 3.987, 4, 10.34566)
round(y, 2)
```

## Getting help in R

### The help() function

Help on a given function can be obtained using the `help()` function. Help can also be called up using the "?" character. Help sheets about functions provide information on input arguments, returned values and usage through examples.

The various help fields are as follows:

- **Description:** provides information on the function's role.
- **Usage:** describes the function and its default parameters.
- **Arguments:** lists the arguments to be passed to this function.
- **Value:** the type of object(s) returned by the function.
- **See Also:** other similar or complementary functions.
- **Examples:** some examples of how to use this function}.

<div class="exo">
- Look for help on one of this functions: `min()`, `max()`, `sort()`, `log()`, `log2()` or `log10()`.

```{r function6, exercise=TRUE, exercise.setup = "function1"}

```

```{r function6-solution}
?min
# or
help(min)
```

</div>

## Exercises

### Exercise 1

Answer the following questions:

<div class="exo">
```{r exoq1, echo=FALSE }
learnr::question("It's advisable to use the '=' operator rather than '<-' when assigning ?",
  answer("Yes"),
  answer("No", correct = TRUE),
  allow_retry = TRUE 
)
```

```{r exoq2, echo=FALSE }
learnr::question("When calling a function, if the arguments are named, their order doesn't matter.",
  answer("Yes", correct = TRUE),
  answer("No"),
  allow_retry = TRUE 
)
```

```{r exoq4, echo=FALSE }
learnr::question("When calling a function, if you abbreviate the names of the arguments, take care that the abbreviations are not ambiguous.",
  answer("Yes", correct = TRUE),
  answer("No"),
  allow_retry = TRUE 
)
```

```{r exoq3, echo=FALSE }
learnr::question("The ls() function is used to list files?",
  answer("Yes"),
  answer("No", correct = TRUE),
  allow_retry = TRUE 
)
```

</div>

### Finished

Thank you for following this tutorial.
